---
title: "Class 11 Structural Bioinformatics"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# What is in the PDB database

Download PDB statistics summary sorted by experimental method and molecular type

```{r}
data <- read.csv("Data Export Summary.csv")
```

Q. Determine the percentage of structures solved by X-Ray and Electron Microscopy
 
```{r}
sum(data$Total)
```

```{r}
ans <- data$Total/sum(data$Total)*100
names(ans) <- data$Experimental.Method
ans
round(ans,2)
```

Q. What proportion of structures are protein

```{r}
round(sum(data$Proteins)/sum(data$Total)*100, 2)
```

Q2: Type HIV in the PDB website search box on the home page and determine how many HIV-1 protease structures are in the current PDB?
3048

# WOrking with biomolecular data in R

```{r}
library(bio3d)
pdb <- read.pdb("1hsg")
pdb
```

```{r}
pdb$atom[1,"resid"]
```

```{r}
aa321(pdb$atom[pdb$calpha,"resid"])
```

```{r}
# Select all C-alpha atoms (return their indices)
ca.inds <- atom.select(pdb, "calpha")
ca.inds
```

```{r}
# Print details of the first few selected atoms
head(pdb$atom[ca.inds$atom,])
```

```{r}
# Selected xyz coordinates
head(pdb$xyz[, ca.inds$xyz])
```

Q8. Use the Bio3D write.pdb() function to write out a protein only PDB file for viewing in VMD. Also write out a second separate PDB file for the ligand with residue name MK1

First select a protein and write out a file "1hsg_protein.pdb"
```{r}
prot <- atom.select(pdb, "protein", value=TRUE)
write.pdb(prot, file="1hsg_protein.pdb")
```

Select ligand and write out a file
```{r}
lig <- atom.select(pdb, "ligand", value=TRUE)
write.pdb(lig, file="1hsg_ligand.pdb")
```

```{r}
# The 'devtools' package allows us to install development versions
#install.packages("devtools")
# Install the bio3d.view package from bitbucket
#devtools::install_bitbucket("Grantlab/bio3d-view")
```

```{r}
# Load the package
library("bio3d.view")
# view the 3D structure
view(pdb, "overview", col="sse")
```

```{r}
# Load the package
pdb <- read.pdb("1hel")
# Normal mode analysis calculation
modes <- nma(pdb)
m7 <- mktrj(modes,
 mode=7,
 file="mode_7.pdb")
view(m7, col=vec2color( rmsf(m7) ))

```

