---
title: "Class 8 Machine Learning 1"
author: "Fabian Lim"
date: "10/25/2019"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## K-means clusterng in R
```{r}
# Generate some example data for clustering
tmp <- c(rnorm(30,-3), rnorm(30,3))
x <- cbind(x=tmp, y=rev(tmp))
plot(x)
```

Use the kmeans() function setting k to 2 and nstart=20

```{r}
k <- kmeans (x, centers = 2, nstart =20)
```

Inspect/print the results
```{r}
k
```

Q. How many points are in each cluster?
```{r}
k$size
```

Q. What ‘component’ of your result object details
 - cluster size?
 - cluster assignment/membership?
 - cluster center?
```{r}
k$cluster
```

 
Plot x colored by the kmeans cluster assignment and
 add cluster centers as blue points
 
```{r}
plot(x, col=k$cluster)
points(k$centers, col="blue", pch=15)
```
 
 ## Hierarchical clustering in R
```{r}
hc <- hclust(dist(x))
hc
```
 
 Plot my results
```{r}
plot(hc)
abline(h=6, col="red")
abline(h=4, col="blue")
#cutree(hc, h=3)
```
 
```{r}
grps <- cutree(hc, k=2)
table(grps)
```
 
 I can also cut the tree to yield a given k groups/ clusters
 
```{r}
cutree(hc, k=2)
```
 
```{r}
plot(x, col=grps)
```
 
```{r}
# Step 1. Generate some example data for clustering
x <- rbind(
 matrix(rnorm(100, mean=0, sd = 0.3), ncol = 2), # c1
 matrix(rnorm(100, mean = 1, sd = 0.3), ncol = 2), # c2
 matrix(c(rnorm(50, mean = 1, sd = 0.3), # c3
 rnorm(50, mean = 0, sd = 0.3)), ncol = 2))
colnames(x) <- c("x", "y")
# Step 2. Plot the data without clustering
plot(x)
# Step 3. Generate colors for known clusters
# (just so we can compare to hclust results)
col <- as.factor( rep(c("c1","c2","c3"), each=50) )
plot(x, col=col)
```

 Q. Use the dist(), hclust(), plot() and cutree()
 functions to return 2 and 3 clusters 
```{r}
hc <- hclust(dist(x))
plot(hc)
```
 
```{r}
cl2 <- cutree(hc, k=2)
plot(x, col=cl2)
```
 
 
```{r}
cl3 <- cutree(hc, k=3)
plot(x, col=cl3)
```
 
```{r}
table(cl3)
```
 
 
```{r}
table(col, cl3)
```
 
 ## PCA
 
```{r}
mydata <- read.csv("https://tinyurl.com/expression-CSV",
 row.names=1)
head(mydata)

mydata2 <- read.csv("https://tinyurl.com/expression-CSV")
```
 
```{r}
nrow(mydata) #number of samples
ncol(mydata) #number of genes
```
 
```{r}
## lets do PCA
pca <- prcomp(t(mydata), scale=TRUE)
## See what is returned by the prcomp() function
attributes(pca)
```
```{r}
pca$x
```
 
```{r}
pca$x[,1] #column 1 of pca
```
 
```{r}
plot(pca$x[,1], pca$x[,2])
```
 
```{r}
summary
#cumultive proportion is the cumulation of the pca components
```
 
```{r}
pca.var <- pca$sdev^2
pca.var.per <- round(pca.var/sum(pca.var)*100, 1)

barplot(pca.var.per, main="Scree Plot",
 xlab="Principal Component", ylab="Percent Variation")
```
 
```{r}
## A vector of colors for wt and ko samples
colvec <- as.factor( substr( colnames(mydata), 1, 2) )
plot(pca$x[,1], pca$x[,2], col=colvec, pch=16,
 xlab=paste0("PC1 (", pca.var.per[1], "%)"),
 ylab=paste0("PC2 (", pca.var.per[2], "%)")) 
```

 ## PCA on UK foods database

```{r}
x <- read.csv("UK_foods.csv", row.names=1)
```
 
```{r}
nrow(x)
ncol(x)
```
 
```{r}
head(x)
```
 
```{r}
dim(x) #dimensions
```
 
```{r}
barplot(as.matrix(x), beside=F, col=rainbow(nrow(x)))
```
 
```{r}
pairs(x, col=rainbow(10), pch=16)
```
 N. Ireland has an outlier consistently when plot against all other countries
 
```{r}
# Use the prcomp() PCA function 
pca <- prcomp( t(x) )
summary(pca)
```
 
```{r}
# Plot PC1 vs PC2
plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2", xlim=c(-270,500), col=c("yellow", "red", "blue", "green"))
text(pca$x[,1], pca$x[,2], colnames(x), col=c("yellow", "red", "blue", "green"))
```
 
 
 