---
title: "Class 14"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



##Setup Bioconductor DESeq2

```{r eval=FALSE}
install.packages("BiocManager")
BiocManager::install()
n to update all/some/none
BiocManager::install("DESeq2")
```

## Data for todays class

```{r}
counts <- read.csv("airway_scaledcounts.csv", stringsAsFactors=F)
metadata <- read.csv("airway_metadata.csv", stringsAsFactors = F)
```

Lets have a peak..
```{r}
head(counts)
```

How many genes do we have in this dataset?
```{r}
nrow(counts)
```

How many experiments
```{r}
ncol(counts)-1
```

Lets make sure metadata id col matches the colnames of counts
```{r}
all(colnames(counts)[-1] == metadata$id)
```

The all funciton is useful for looking across a vector of logicals
```{r}
all(c(F,T,T))
all(c(T,T,T))
```

## Analysis: compare the control to drug treated

First we need to access the columns of our countData that are control and ten find their mean

```{r}
metadata$dex=="control"
```

```{r}
control.id <- metadata[metadata$dex=="control", ]$id
control.id
```

```{r}
head(counts[,control.id])
```

Mean counts for control exps


```{r}
control.mean <- rowSums(counts[,control.id])/length(control.id)
names(control.mean) <- counts$ensgene
```


## Treated samples

```{r}
metadata$dex=="treated"
```

```{r}
treated.id <- metadata[metadata$dex=="treated", ]$id
treated.id
```


```{r}
head(counts[,treated.id])
```

Mean counts for treated exps


```{r}
treated.mean <- rowSums(counts[,treated.id])/length(treated.id)
names(treated.mean) <- counts$ensgene
```

Lets store these together

```{r}
meancounts <- data.frame(control.mean, treated.mean)
```

Plot control vs treated

```{r}
plot(meancounts$control.mean, meancounts$treated.mean)
```

Good case for a log plot

```{r}
plot(meancounts$control.mean, meancounts$treated.mean, log="xy")
```


## Fold change

```{r}
meancounts$log2fc <- log2(meancounts[, "treated.mean"]/meancounts[, "control.mean"])
head(meancounts)
```

There are a couple of “weird” results. Namely, the NaN (“not a number”“) and -Inf (negative infinity) results.

The NaN is returned when you divide by zero and try to take the log. The -Inf is returned when you try to take the log of zero. It turns out that there are a lot of genes with zero expression. Let’s filter our data to remove these genes. Again inspect your result (and the intermediate steps) to see if things make sense to you

```{r}
#Test for finding zero entries

x <- c(5,8,0,5)
x==0
which(x==0)
x <-data.frame(happy=c(5,6,0,0), sad=c(0,5,5,0))
x==0
```

```{r}
which(x==0, arr.ind=TRUE)
```

Focus on the rows only that have zero entries

```{r}
unique(which(x==0, arr.ind=TRUE)[,1])
```

Think pair share

```{r}
x <-data.frame(happy=c(6,3,2,0), sad=c(5,0,5,0))
x==0
```

```{r}
which(x==0, arr.ind=TRUE)
```

```{r}
unique(which(x==0, arr.ind=TRUE)[,1])
```

Now do this for our meancounts data, i.e. remove zero containing genes from further consideration

```{r}
to.rm <- unique(which(meancounts[,1:2]==0, arr.ind=TRUE)[,1])
mycounts <- meancounts[-to.rm,]
head(mycounts)
```

A common threshold used for calling something differentially expressed is a log2(FoldChange) of greater than 2 or less than -2. Let’s filter the dataset both ways to see how many genes are up or down-regulated.

```{r}
up.ind <- mycounts$log2fc > 2
down.ind <- mycounts$log2fc < (-2)
```

```{r}
sum(up.ind)
```

```{r}
sum(down.ind)
```

```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")
```

See what is available in org.Hs

```{r}
columns(org.Hs.eg.db)
```

```{r}
mycounts$symbol <- mapIds(org.Hs.eg.db,
                     keys=row.names(mycounts),
                     column="SYMBOL",
                     keytype="ENSEMBL",
                     multiVals="first")
```

```{r}
head(mycounts)
```

# DESeq2 analysis

```{r}
library(DESeq2)
```

```{r}
dds <- DESeqDataSetFromMatrix(countData=counts, 
                              colData=metadata, 
                              design=~dex, 
                              tidy=TRUE)
dds
```

```{r}
dds <- DESeq(dds)
```

```{r}
res <- results(dds)
res
```

# Volcano plot

```{r}
plot(res$log2FoldChange, res$padj)
```

```{r}
mycols <- rep("gray", length(res$padj))
mycols[ abs(res$log2FoldChange) >2] = "blue"
mycols[(res$padj > 0.05) & (abs(res$log2FoldChange) >2)] = "red"
plot(res$log2FoldChange, -log(res$padj), col=mycols)
```



